Highly Available Transactions: Virtues and Limitations

### 2013-03-29 writeup/evaluation.tex

TPC-C consists of five transactions, capturing operation of a
real-world warehouse, including sales, payments, and deliveries. Two
transactions---Order-Status and Stock-Level---are read-only and can be
executed safely under HATs. Clients may read stale data, but this does
not violate TPC-C requirements and clients will read their writes if
they are sticky-available. Another transaction type, Payment, updates
running balances for warehouses, districts, and customer records as
well as providing an audit trail. The transaction is increment-only,
so all balance increase operations can commute, and TA allows the
maintainence of any foreign-key constraints.

@@

>>>>

TPC-C consists of five transactions, capturing operation of a
real-world warehouse, including sales, payments, and deliveries. Two
transactions---Order-Status and Stock-Level---are read-only and can be
executed safely under HATs. Clients may read stale data, but this does
not violate TPC-C requirements and clients will read their writes if
they are sticky-available. Another transaction type, Payment, updates
running balances for warehouses, districts, and customer records as
well as providing an audit trail. The transaction is increment-only,
so all balance increase operations can commute, and TA allows the
maintainence of any foreign-key constraints (e.g.,
\texttt{UPDATE/DELETE CASCADE}).

@@

\begin{figure}
\includegraphics[width=.8\columnwidth]{figs/macrotracelegend.pdf}\vspace{-2mm}
\includegraphics[width=\columnwidth]{figs/lat-thru-trace-K0-SCALEMACRO.pdf}
\includegraphics[width=\columnwidth]{figs/lat-thru-trace-K0-SCALEMACRO.pdf}
\caption{Latency-throughput for database operation on EC2 (DUMMY FIG)}
\label{fig:wan-exp}
\end{figure}

### 2013-03-29 writeup/hat-analysis.tex

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the acronym, it is not an ``isolation'' property,
transactional atomicity restricts transactions' ability to view the
effects of partially completed transactions. Under transactional
atomicity, within each transaction, either all effects of another
transaction are observed, or none are (equivalently, once some of the
effects of a transaction are observed, all effects are observed). This
is a strictly stronger guarantee than Read Committed isolation: in
Read Committed, we can read a subset of committed writes whereas, with
transactional atomicity, we must read all of them. Together with cut
isolation, transactional atomicity prevents Read Skew anomalies
(Berenson et al.'s A5A~\cite{ansicritique}). As an example of
TA, $T_2$ must observe $b=c=1$ (or later versions
for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$ or $a=1$.

@@

The inability to prevent Lost Update means that Consistent Read,
Snapshot Isolation, and Cursor Stability guarantees are all
unavailable. The inability to prevent Lost Update or Write Skew means
that Repeatable Read, Conflict Serializability, and One-Copy
Serializability guarantees are unavailable.

@@

We illustrate the hierarchy of available, sticky available, and
unavailble consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable: for example, a HAT system can provide monotonic reads and
transactional atomicity or transactional atomicity and read your
writes and, for maximum semantic richness, can provide all of
transactional atomicity, predicate cut isolation, monotonic reads,
monotonic writes, and writes follow reads. An unavailable system might
provide PRAM and Serializability, as advocated by Salem and
Daudjee~\cite{daudjee-session}.

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Perhaps satisfactorily, strong one-copy serializability
subsumes all other models, while considering the (large) power set of
all compatible models (e.g., the diagram depicts 144 possible highly available
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonimization is \textit{not} exhaustive, but we
believe it lends substantial clarity into the relationships between a
large subset of the prominent ACID and distributed consistency
moels. Additional models of read/write transactions that we have
omitted should be easily classifiable based on the available
primitives and unavailable anomalies we have already discussed.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (P-CI), Predicate Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, Conflict Serializability
(CSR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Strong 1SR\lostupdate\rwskew\linearizable
\\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \node[draw=none,fill=none] (ici) at (0, 0) {I-CI};
 \node[draw=none,fill=none] (pci) at (1.1, 1.3) {P-CI};
 \node[draw=none,fill=none] (rc) at (-1.2, .8) {RC};
 \node[draw=none,fill=none] (ru) at (-1.2, 0) {RU};

 \node[draw=none,fill=none] (ta) at (1.2, 0) {TA};
 \node[draw=none,fill=none] (mr) at (2.4, 0) {MR};
 \node[draw=none,fill=none] (mw) at (3.6, 0) {MW};
 \node[draw=none,fill=none] (wfr) at (6.3,0) {WFR};
 \node at (5,0) [sticky] (ryw) {RYW};

 \node[noha](recency) at (7.7, 0) {recency};
 \node[noha](safe) at (7.7, 1) {safe};
 \node[noha](regular) at (7.7, 2) {regular};
 \node[noha](linearizable) at (7.7, 3) {linearizable};
 \node at (3.6, 2) [sticky] (causal) {causal};
 \node at (3.6, 1) [sticky] (pram) {PRAM};
 \node[noha] (cs) at (-1.2, 1.8) {CS};
 \node[noha] (rr) at (-.8, 2.7) {RR};
 \node[noha] (si) at (2.2, 2.4) {SI};
 \node[noha] (s) at (0, 3.5) {CSR};
 \node[noha] (sr) at (2.2, 3.6) {1SR};
 \node[noha] (ssr) at (4.95, 3.7) {Strong-1SR};

 \draw [->, red] (recency) -- (safe);
 \draw [->, red] (safe) -- (regular);
 \draw [->, red] (regular) -- (linearizable);
 \draw [->, red] (linearizable) -- (ssr);
 \draw [->, red] (sr) -- (ssr);

 \draw [->] (ru) -- (rc);
 \draw [->] (ici) -- (pci);

@@

 \draw [->, red] (rc) -- (cs);
 \draw [->, red] (cs) -- (rr);
 \draw [->, red] (pci) -- (si);
 \draw [->, red] (ici) -- (rr);
 \draw [->, red] (rr) -- (s);
 \draw [->, red] (si) -- (s);
 \draw [->, red] (ta) -- (si);
 \draw [->, red] (ta) -- (rr);
 \draw [->, red] (s) -- (sr);
 \draw [->, red] (causal) -- (sr);

>>>>

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the acronym, it is not an ``isolation'' property,
transactional atomicity restricts transactions' ability to view the
effects of partially completed transactions. Under transactional
atomicity, within each transaction, either all effects of another
transaction are observed, or none are (equivalently, once some of the
effects of a transaction are observed, all effects are observed). This
is a strictly stronger guarantee than Read Committed isolation: in
Read Committed, we can read a subset of committed writes whereas, with
transactional atomicity, we must read all of them. Together with cut
isolation, transactional atomicity prevents Read Skew anomalies
(Berenson et al.'s A5A~\cite{ansicritique}). As an example of
TA, $T_2$ must observe $b=c=1$ (or later versions
for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$ or $a=1$. Notably, TA implies Read
Committed isolation: observing all effects of a transaction implicitly
requires observing the final (committed) effects of a transaction as
well.

@@

The inability to prevent Lost Update means that Consistent Read,
Snapshot Isolation, and Cursor Stability guarantees are all
unavailable. The inability to prevent Lost Update or Write Skew means
that Repeatable Read and One-Copy Serializability guarantees are
unavailable.

@@

We illustrate the hierarchy of available, sticky available, and
unavailble consistency models we have discussed in
Figure~\ref{fig:hatcompared}. Many models are simultaneously
achievable, but we find several particularly compelling. If we combine
all sticky-HAT guarantees, we have transactional, causal snapshot
reads. If we combine TA and P-CI, we have transactional snapshot
reads. We can achieve RC, MR, and RYW by simply sticking clients to
servers. And we can also combine unavailable models---for example, an
unavailable system might provide PRAM and One-Copy
Serializability~\cite{daudjee-session}.

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Perhaps satisfactorily, strong one-copy serializability
subsumes all other models, while considering the (large) power set of
all compatible models (e.g., the diagram depicts 96 possible highly available
combinations) hints at the vast expanse of consistency models found in
the literature. This taxonimization is \textit{not} exhaustive, but we
believe it lends substantial clarity into the relationships between a
large subset of the prominent ACID and distributed consistency
moels. Additional models of read/write transactions that we have
omitted should be easily classifiable based on the available
primitives and unavailable anomalies we have already discussed.

@@

\begin{table}[t!]
\begin{tabular}{| c | p{6cm} | }\hline
HA & Transactional Atomicity (TA), Read Uncommitted (RU), Read
Committed (RC), Item Cut Isolation (P-CI), Predicate Cut Isolation
(P-CI), Monotonic Reads (MR), Monotonic Writes (MW), Writes Follow
Reads (WFR)\\\hline Sticky-HA & Read Your Writes (RYW), PRAM,
Causal\\\hline Unavailable & Cursor Stability (CS)\lostupdate,
Snapshot Isolation (SI)\lostupdate, Repeatable Read
(RR)\lostupdate\rwskew, One-Copy Serializability
(1SR)\lostupdate\rwskew, Recency\linearizable, Safe\linearizable,
Regular\linearizable, Strong 1SR\lostupdate\rwskew\linearizable
\\\hline
\end{tabular}
\caption{Summary of highly available, sticky highly available, and
  unavailable models considered in this paper. Unavailable models are
  labeled by cause of unavailability: preventing lost
  update\lostupdate, preventing write skew\rwskew, and requiring
  recency guarantees\linearizable.}
\label{table:hatcompared}
\end{table}

@@

 \node[draw=none,fill=none] (ici) at (1.2, 0) {I-CI};
 \node[draw=none,fill=none] (pci) at (1.65, 1.2) {P-CI};
 \node[draw=none,fill=none] (rc) at (-1.2, .8) {RC};
 \node[draw=none,fill=none] (ru) at (-1.2, 0) {RU};

 \node[draw=none,fill=none] (ta) at (-.2, 1.3) {TA};
 \node[draw=none,fill=none] (mr) at (2.4, 0) {MR};
 \node[draw=none,fill=none] (mw) at (3.6, 0) {MW};
 \node[draw=none,fill=none] (wfr) at (6.3,0) {WFR};
 \node at (5,0) [sticky] (ryw) {RYW};

 \node[noha](recency) at (7.7, 0) {recency};
 \node[noha](safe) at (7.7, 1) {safe};
 \node[noha](regular) at (7.7, 2) {regular};
 \node[noha](linearizable) at (7.7, 3) {linearizable};
 \node at (3.6, 2) [sticky] (causal) {causal};
 \node at (3.6, 1) [sticky] (pram) {PRAM};
 \node[noha] (cs) at (-1.2, 1.8) {CS};
 \node[noha] (rr) at (-.2, 2.7) {RR};
 \node[noha] (si) at (2.2, 2.4) {SI};
 \node[noha] (1sr) at (1.2, 3.2) {1SR};
 \node[noha] (ssr) at (3.6, 3.6) {Strong-1SR};

 \draw [->, red] (recency) -- (safe);
 \draw [->, red] (safe) -- (regular);
 \draw [->, red] (regular) -- (linearizable);
 \draw [->, red] (linearizable) -- (ssr);
 \draw [->, red] (1sr) -- (ssr);

 \draw [->] (ru) -- (rc);
 \draw [->] (rc) -- (ta);
 \draw [->] (ici) -- (pci);

@@

 \draw [->, red] (rc) -- (cs);
 \draw [->, red] (cs) -- (rr);
 \draw [->, red] (pci) -- (si);
 \draw [->, red] (ici) -- (rr);
 \draw [->, red] (rr) -- (1sr);
 \draw [->, red] (si) -- (1sr);
 \draw [->, red] (ta) -- (si);
 \draw [->, red] (ta) -- (rr);
 \draw [->, red] (causal) -- (ssr);

### 2013-03-29 writeup/hat-vldb.tex
draftwatermark
Another pass at writing
more writing

\documentclass{vldb}
\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)

@@

@@

\title{Highly Available Transactions: Virtues and Limitations}

\maketitle

@@

foo~\cite{adya}

>>>>

\documentclass{vldb}
\usepackage{times, graphicx, url, xcolor, multicol, minipage, subfig, tikz, draftwatermarkbottom, draftwatermarktop}
\usepackage{balance}  % for  \balance command ON LAST PAGE  (only there!)

\renewcommand{\ttdefault}{fi4}

@@

\newenvironment{myitemize}
{

   \vspace{0mm}
    \begin{list}{$\bullet$}{\leftmargin=1em}
        \setlength{\topsep}{0em}
        \setlength{\parskip}{0pt}
        \setlength{\partopsep}{0pt}
        \setlength{\parsep}{0pt}         
        \setlength{\itemsep}{.25em} 
        \setlength{\itemindent}{0em}
}
{
    \end{list} 
    \vspace{-.5em}
}

\newenvironment{myenumerate}
{

   \vspace{0mm}
   \newcounter{qcounter}
    \begin{list}{\arabic{qcounter}.~}{\usecounter{qcounter}\leftmargin=1em}
        \setlength{\topsep}{0em}
        \setlength{\parskip}{0pt}
        \setlength{\partopsep}{0pt}
        \setlength{\parsep}{0pt}         
        \setlength{\itemsep}{.25em} 
        \setlength{\itemindent}{0em}
}
{
    \end{list} 
    \vspace{-.5em}
}

@@

\title{Highly Available Transactions: Virtues and Limitations}
{\author{Peter Bailis{\fontsize{12}{14}$^\dagger$}, Aaron Davidson{\fontsize{12}{14}$^\dagger$}, Alan Fekete{\fontsize{12}{14}$^\diamond$}, Ali Ghodsi{\fontsize{12}{14}$^{\dagger,\ddagger}$}, Joseph M. Hellerstein{\fontsize{12}{14}$^\dagger$}, Ion Stoica{\fontsize{12}{14}$^\dagger$}\\{{\fontsize{12}{14}$^\dagger$}\hspace{.5mm}UC Berkeley\hspace{4mm}{\fontsize{12}{14}$^\diamond$}\hspace{.5mm}University of Sydney\hspace{4mm}{\fontsize{12}{14}$^\ddagger$}\hspace{.5mm}KTH/Royal Institute of Technology}}}
\maketitle

@@

### 2013-03-29 writeup/relatedwork.tex
text editing pass

VI. Related Work (.75pg)

* Probably do a good job with this one, in addition to the first part of Section V. We can talk at length about the relationship between distributed systems and databases.

1.) High Availability and CAP (giving the DB folks due credit here!)
2.) Highly Available Distributed Systems
3.) Weak Isolation (there's a lot here, pay plenty of lip service to Adya)
4.) Available/unavailable systems
	--Stress that, with the advent of Spanner, transactions are coming back; HATS represents new design space. (Can even say "Spanner borrowed traditional database techniques; would benefit from HAT approach (c.f., 100ms geo-replicated latencies)"; also talk about optimization for read-only transactions).

>>>>

Although we are not aware of any related work specifically addressing
what properties are available in a highly available system, there are
several systems and related research results upon which we draw in
this work.

Weak consistency and high availability have been well
studied. Serializability has long been known to be
unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
attracted considerable attention~\cite{gilbert-cap}. The research has
begun to offer hints regarding the connection between ``weak
consistency'' and low latency~\cite{abadi-pacelc}, while several
studies~\cite{adya, ansicritique} examine weak isolation guarantees,
grounded in a single-server context. The connection between weak
consistency models in a distributed setting~\cite{calm, crdt} and high
availability is relatively unexplored. With the exception of a
preliminary workshop paper on this work~\cite{hat-hotos}---upon which
this work builds, providing impossibility results, a full
taxonomization, experimental and application analysis (basically all
but Section 4.1)---we believe this paper is the first to explore this
area.

There has been a recent resurgence of interest in distributed
multi-object semantics, both in academia~\cite{kraska-s3, granola,
  gstore, mdcc, redblue, cops, eiger, walter,calvin, swift} and
industry~\cite{megastore, orleans, spanner}. We can roughly categorize
related work into two categories: unavailable systems with stronger
semantics than HATs and highly available systems with weaker
semantics.

Unavailable, strongly consistent databases have a long tradition. As
discussed in Section~\ref{sec:motivation}, classic ACID databases
provide strong semantics but their lock-based and traditional
multi-versioned implementations are unavailable in the presence of
partitions~\cite{bernstein-concurrency, bernstein-book,
  gray-isolation}. Similar semantics are common in recent
work. Notably, Google's Spanner provides externally serializable
transactions. While Spanner fits Google's read-mostly workload well,
read/write transactions use two-phase locking (incurring at least one
round trip time over wide-area networks), are unavailable to clients
in a non-majority partition, and cannot read their
writes~\cite{spanner}. These latency and availability penalties are
fundamental to providing strong consistency. For users willing to
tolerate these costs, Spanner, or similar strongly consistent,
unavailable systems---including Calvin~\cite{calvin},
G-Store~\cite{gstore}, Gemini~\cite{redblue}, Granola~\cite{granola},
HBase~\cite{hbase}, MDCC~\cite{mdcc}, Megastore~\cite{megastore},
Orleans~\cite{orleans}, and Walter~\cite{walter}---are a reasonable
choice. With HATs, we seek an alternative set of semantics that are
still useful but do not violate requirements for high availability or
low latency.
